---
title: "pi-squared-example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pi-squared-example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Setup
```{r setup}
library(pisquared)
require(rstan)
require(doMC)
registerDoMC(2)
```

First we'll simulate some p-values with specified non-null rates and sharing level. We'll use `Beta(a=0.2,b=2)` for the non-null pvalues: 
```{r nonnull-p-distribution, fig.width=5, fig.height=4}
true_alpha = 0.2
true_beta = 2
plot(function(p) dbeta(p, true_alpha, true_beta), ylim = c(0,5), xlab = "pvalue", ylab = "pdf")
```

We use pisquared's built-in function to simulate p-values with specified non-null rates and sharing. Here we set a low level of sharing. Note that some combinations of `pi_1`, `pi_2` and `true_jaccard` are not feasible, in which case you'll get a warning and `NULL` will be returned. 
```{r simulate-low-sharing, fig.width=5, fig.height=4}
set.seed(12345)
dat = simulate_joint_pvalues(N = 10000, # typical num hypothesis tests in genomics
                            pi_1 = 0.1, # marginal pi1 (proportion non-null) for set 1
                            pi_2 = 0.1, # marginal pi1 (proportion non-null) for set 2
                            true_jaccard = 0.1, # true sharing level
                            alpha = true_alpha, # parameters of the beta distribution on 
                            beta = true_beta)
plot(-log10(dat$p1), -log10(dat$p2))
```

Let's see if pisquared can recover the sharing level and marginal pi1: 
```{r estimate-low-sharing}
pi2_results = pi2_estimator(dat$p1, dat$p2)
pi2_results$jaccard
```
This is pretty close to the true level of 0.1. 

What about for higher sharing level of 0.9? 
```{r simulate-high-sharing, fig.width=5, fig.height=4}
set.seed(12345)
dat = simulate_joint_pvalues(N = 10000, # typical num hypothesis tests in genomics
                            pi_1 = 0.1, # marginal pi1 (proportion non-null) for set 1
                            pi_2 = 0.1, # marginal pi1 (proportion non-null) for set 2
                            true_jaccard = 0.9, # true sharing level
                            alpha = true_alpha, # parameters of the beta distribution on 
                            beta = true_beta)
plot(-log10(dat$p1), -log10(dat$p2))
```

It's pretty tough to test by eye whether there is much overlap between the tests. 
```{r estimate-high-sharing}
pi2_results = pi2_estimator(dat$p1, dat$p2)
pi2_results$jaccard
```
This is pretty close to the true level of 0.9. 

For more intensive investigation of the actual of pisquared across a range of sharing levels see `vignette("pisquared-extended")`. 
